\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-02-21 paper/si.tex
working on paper in folder paper

\definecolor{darkblue}{rgb}{0,0,.8}
\definecolor{darkgreen}{rgb}{0,0.5,0.1}
\newcommand{\matlabfunc}[1]{\textcolor{darkblue}{#1}}
\newcommand{\matlabcomment}[1]{\textcolor{darkgreen}{#1}}
\newcommand{\christian}[1]{\textcolor{blue}{\textbf{CH}: #1}}
\newcommand{\alex}[1]{\textcolor{red}{\textbf{AL}: #1}}

@@

\title{\bf  \sffamily \LARGE Supplementary Information: Evolution of cooperation among individuals with
limited payoff memory\\}
\date{}
\author{Nikoleta E. Glynatsi, Christian Hilbe, Alex McAvoy}

@@

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. Section~\ref{section:updating_payoffs} describes the
conventional approach for calculating updating payoffs, as well as, our
approaches and methods.

@@

Pairwise comparison process is a stochastic process for modelling the evolution
of a finite population. The process starts with assigning all individuals of the
population the same strategy. A strategy is a set of rules of how an individual
should behave in an interaction with another individual. Each elementary time
step of the process consists of three phases; (1) the \textbf{mutation phase}
(2) the \textbf{game phase} (3) the \textbf{update phase}. These are summarised
in Figure~\ref{fig:pairwise_phases}. In the mutation phase one individual is
chosen to switch to a new mutant strategy with a probability \(\mu\). The mutant
strategy is randomly selected from the set of feasible strategies.

In the game phase individuals are randomly matched with other individuals in the
population, and they engage in a match where each subsequent turn occurs with a
fixed probability \(\delta\). At each turn the individuals play a social dilemma
and decide on an action based on their strategies. In the donation game there
are two actions: cooperation (\(C\)) and defection (\(D\)). By cooperating a player
provides a benefit \(b\) to the other player at their cost \(c\), with \(0 < c <
b\). Thus the payoffs for a player in each turn are,

Let \(u = (b-c, -c, b, 0)\) be payoffs in a vector format, and let \(\mathcal{U}
= \{b-c, -c, b, 0\}\) denote the set of feasible payoffs. In repeated games
there are infinite many strategies. However, it is commonly assumed that
individuals can only choose strategies from a restricted set. Here we explore
the case where individuals can use reactive strategies. A reactive strategy
considers only the previous action of the other player, and thus, a reactive
strategy \(s\) can be written as a three-dimensional vector \(s=(y, p, q)\). The
parameter \(y\) is the probability that the strategy opens with a cooperation
and \(p\), \(q\) are the probabilities that the strategy cooperates given that
the opponent cooperated and defected equivalently.

Following the game phase is the update stage where two individuals are randomly selected.
One individual serves as the `learner' and the other as the `role model'.
The learner adopts the role model's strategy with a probability \(\rho\) given by,

\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{RM}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (\pi_{RM}\!-\! \pi_{L})}}.
\end{equation}

\(\pi_{L}\) and \(\pi_{RM}\) are the updating payoffs of the learner and the
role model respectively. The updating payoff is a measure of how successful an
individual is in the current standing of the population. In the next section we
will explain how these payoffs are calculated in more detail. The parameter
\(\beta\) is known as the selection strength, namely, it shows how important the
payoff difference is when the learner is considering adopting the strategy of
the role model.

\begin{figure}[!htbp]
    \centering
    \includestandalone[width=\textwidth]{pairwise_comparison_process_diagram}
    \caption{\textbf{Pairwise process phases.} \textbf{A)} The process begins
    with a finite population where each member is assigned a given strategy.
    Each color represents a different strategy, and the members are labelled by
    letters. \textbf{B) Mutation phase.} An individual is selected (in the
    example individual A) and with a given probability \(\mu\) that individual
    adopts a new mutant strategy. \textbf{C) Game phase.} individuals are
    selected to interact in a repeated social dilemma with other individuals. We
    demonstrate the case where individuals B and F have been selected to
    interact. They use the reactive strategies \(s_{B} = (y_B, p_B, q_B)\) and
    \(s_{F} = (y_F, p_F, q_F)\) respectively. The opening moves depend on their
    \(y_i\) probability. In turn 1, individual B cooperated,
    thus, F cooperates with a probability \(p_{F}\) in turn 2. On the opposite,
    individual F defected in turn 1, and so B cooperates in the next turn with
    a probability \(q_{B}\). At each turn there is a probability \(delta\) that a
    subsequent turn will occur, and with a probability \(1- delta\) the interaction ends.
    \textbf{D) Updating phase.} At the updating phase two individuals are chosen;
    one is assigned the role of the learner and the other one the role
    of the role model. In our example C adopts D's strategy with a probability
    \(\rho(\pi_{C}, \pi_{D})\) where \(pi_{C}, \pi_{D}\) are the updating
    payoffs of the individuals.}\label{fig:pairwise_phases}
\end{figure}

This elementary population updating process is repeated for a large number of
time steps, and at each time step we record the state of the population.

@@

\begin{algorithm}[!htbp]
  \SetAlgoLined $N \leftarrow$ population size\; $k \leftarrow 1$\; resident
   $\leftarrow (0, 0, 0)$\; \While{ $t <$ maximum number of steps}{mutant
   $\leftarrow$ random: $\{\emptyset \}\ \rightarrow R^3$\; fixation probability
   $\leftarrow \varphi $\; \If{$\varphi >$ random: $i \rightarrow [0,
   1]$}{resident $\leftarrow$ mutant;}} \caption{Evolutionary
   process}\label{algorithm:pairwise_comparison}
\end{algorithm}

Most of the results we present in this work consider the case of low mutation,
however, we have also verified that the main result holds even in the case where
\(\mu\) is not small. %ToDo Add reference to section with mu != 0.

@@

In this section we discuss the conventional approach of calculating the updating
payoffs, as well as our newly introduced approach.

@@

The expected payoffs are the conventional payoffs used in the updating stage.
The expected payoffs are defined as the mean payoff of an individual in a
well-mixed population that engages in repeated games with all other population
members. We first define the payoff of two reactive strategies at the game
stage. Assume two reactive strategies $s_1\!=\!(y_1, p_1, q_1$) and
$s_2\!=\!(y_2,p_2,q_2)$. The play between the two strategies can be defined as a
Markov process with the transition matrix \(M\),

\begin{equation}\label{eq:transition_matrix}
  \input{../static/matrix.tex}.
\end{equation}

whose stationary vector \(\mathbf{v}\), combined with the payoff \(u\), yields
the game stage outcome for each strategy,

\[\langle\mathbf{v}(s_1,s_2),\mathbf{u}\rangle \text{ and } \langle\mathbf{v}(s_2,s_1), \mathbf{u}\rangle\].

In the case of low mutation there can be only two different
strategies in the population; a resident and a mutant strategy. Thus, we
only need to define the expected
payoff for a resident (\(\pi_R\)) and for a mutant (\(\pi_M\)). Assume the
resident strategy \(s_R = (y_R, p_R, q_R)\) and the mutant strategy \(s_M =
(y_M, p_M, q_M)\). The expected payoffs are give by,

@@

As a reminder \(N\) is the population size and \(k\) the number of mutants. The
number of mutants in the population increases if a resident adopts the strategy
of a mutant, and decreases if a mutant adopts the strategy of a resident. The
probabilities that the number of mutants decreases and increases,
\(\lambda^-_k\) and \(\lambda^+_k\), are now explicitly defined as,

\begin{align*} 
  \lambda^-_k &\!=\!\rho(\pi_M, \pi_R) \\
  \lambda^+_k &\!=\!\rho(\pi_R, \pi_M).
\end{align*}

\subsection{Invasion analysis of ALLD into Generous Tit For Tat (GTFT)}

In the following, we apply the above formalism to calculate how easily a single
ALLD mutant can invade into a resident population with strategy GTFT. In that
case, \(s_1 = (1, 1, q)\), \(s_2 = (0, 0, 0)\), and \(k = 1\). When two GTFT
players interact in the game, their respective probabilities for each of the
four outcomes in the last round simplify to,

\begin{align*}
    \mathbf{v}(s_1, s_1) = (1, 0, 0, 0).
\end{align*}

@@

\begin{align*}
  \mathbf{v}(s_2, s_1) = (0, q, 0, (1 - q)).
\end{align*}

@@

\begin{align*}
  \displaystyle \pi_R & = \displaystyle \frac{N\!-\!2}{N-1}\cdot (b - c) \cdot -	\displaystyle\frac{1}{N-1}\cdot q \cdot c\text{ and } \displaystyle \pi_M  = \displaystyle b \cdot q.
\end{align*}

@@

\begin{align*}
    \frac{\lambda^{+}}{\lambda^{-}} & = \frac{\rho(\pi_R, \pi_M)}{\rho(\pi_M, \pi_R)} \\[0.5cm]
    \frac{\lambda^{+}}{\lambda^{-}} & = \frac{e^{-\beta \left(\frac{(N-2) (b-c)}{N-1} - q \cdot (b + \frac{c}{N-1})\right)}+1}
    {e^{-\beta\left(b \cdot q - \frac{b (N - 2)}{N-1} - \frac{c (N - 2 - q)}{N-1}\right)}+1}
\end{align*}

@@

\subsection{Simulation results}\label{section:simulation_results_exprected}

Figure~\ref{fig:expected_payoffs_results} shows simulation results of the
evolutionary process when individuals update their strategies based on their
expected payoffs. We run two simulations where we differ the benefit of
cooperation. Though, a higher benefit results in a more cooperative population,
the population is still rather cooperative even in the case of low benefit.

In the case of low benefit the resident population consists either of defectors
or conditional cooperators, whereas in the case of high benefit the residents
typically apply a conditional cooperative strategy. Based on the results on
invasion we can show that conditional cooperators that the residents adopt are
of the kind \(p \approx 1\) and \(q \leq 1 - \frac{c}{b}\).

\begin{figure}[!htbp]
    \centering 
    \includegraphics[width=.70\textwidth]{../static/expected_payoffs_donation_game.pdf}
    \caption{\textbf{Evolutionary dynamics under expected payoffs with
    low (left) and high (right) benefit.} We run the simulations for \(T =
    10^7\) time steps. For each time step, we have recorded the current resident
    population \((y, p, q)\). Since in the case of the expected payoffs the
    individuals interact for an infinite number of times \(\delta \rightarrow
    1\), we do not report the players' initial cooperation probability \(y\).
    The graphs show how often the resident population chooses each combination
    \((p, q)\) of conditional cooperation probabilities in the subsequent
    rounds. In both cases players update based on their expected payoffs (A) In
    the case where \(b=3\), the resident population either consists of defectors
    (with \(p \approx q \approx 0\)) or of conditional cooperators for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{3}=0.7\). (B) In the case where
    \(b=10\), the resident population typically applies a strategy for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{10}=0.9\). Parameters: \(N =100, c=1,
    \beta=1\).
    }\label{fig:expected_payoffs_results}
\end{figure}

\subsection{Updating Payoffs based on the one interaction - last round payoffs}

The first approach we introduced in this work is that of the one  interaction -
last round updating payoff. In this case an individual updates his strategy
based on the last payoff they received against the one other member of the
population. Initially, we define the payoff of a reactive strategy in the last
round at the stage game. This is given by
Proposition~\ref{proposition:last_round}.

@@

      v_{R}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1y_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(q_1+r_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(q_2+r_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
      {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times R,\\[1cm]

      v_{S}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1\bar{y}_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(q_1+r_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
      {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times S,\\[1cm]

      v_{T}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1y_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(q_2+r_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
      {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times T,\\[1cm]

      v_{P}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1\bar{y}_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
      {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)} \times P.
      \end{array}
    \end{equation}

In these expressions, we have used the notation $r_i:=p_i\!-\!q_i$,
$\bar{y}_i\!=\!1\!-\!y_i$, $\bar{q}_i:=1\!-\!q_i$, and
$\bar{r}_i:=\bar{p}_i\!-\!\bar{q}_i=-r_i$ for $i\!\in\!\{1,2\}$.
\end{Prop}

Note that in the proposition we use the general notation of the prisoner's
dilemma, to denote that the results apply to \(2 \times 2\) any symmetric game.

@@

In the updating stage we select a mutant and resident to be either the role
model or the learner. Given that they can interact with only one other member of
the population, they can interact either with each other or either can interact
with another resident or with another mutant. Thus, in each updating stage there
are five possible combinations of pairs (Figure~\ref{fig:single_pairs}).

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=.65\textwidth]{../static/last_rounds_matches.tex}
  \caption{\textbf{Possible pairings combination in the updating stage, given
  that individuals interact with only one other member in the population.} At
  each step of the evolutionary process we choose a role model and a learner to
  update the population. We consider the case where both the role model and the
  learner estimate their fitness after interacting with a single member of the
  population. There are five possible pairings at each step. They interact with
  other with a probability \(\frac{1}{N - 1}\), and thus they do not interact
  with other with a probability \(1 - \frac{1}{N - 1}\). In the latter case,
  each of them can interact with either a mutant or a resident. Both of them
  interact with a mutant with a probability $\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and
  both interact with a resident with a probability
  $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last two possible pairings are that
  either of them interacts with a resident whilst the other interacts with a
  mutant, and this happens with a probability
  $\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.}
  \label{fig:single_pairs}
\end{figure}

Given the last round payoff and possible pair combinations for a single
interaction, we define the probability that the respective last round payoffs of
two players \(s_1, s_2\) are given by $u_1$ and $u_2$ as,

@@

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the game stage, which happens with
probability $\frac{1}{(N\!-\!1)}$. In that case, we note that only those payoff
pairs can occur that are feasible in a direct interaction,
$\splitatcommas{(u_1,u_2)\in \mathcal{U}^2_F:=\big\{ (b-c, b-c), (-c, b), (b, -c), (0, 0)
\big\}}$, as represented by the respective indicator function. Otherwise, if the
learner and the role model did not interact directly, we need to distinguish
four different cases, depending on whether the learner was matched with a
resident or a mutant, and depending on whether the role model was matched with a
resident or a mutant.

Given that $N\!-\!k$ players use the resident strategy
$s_{R}\!=\!(y_{R},p_{R},q_{R})$ and that the remaining $k$ players use the
mutant strategy $s_{M}\!=\!(y_{M},p_{M},q_{M})$, the probability that the number
of mutants increases by one in one step of the evolutionary process can be
written as

@@

\subsubsection{Invasion analysis of ALLD into (GTFT)}

Similarly to the previous section, we calculate how easily a single ALLD mutant
can invade into a resident population with strategy GTFT.

When two GTFT players interact in the game, their respective probabilities for
each of the four outcomes in the last round simplify to,

\begin{align*}
    v_R(GTFT,GTFT) = 1, & \quad v_T(GTFT,GTFT) = 0, \\
    v_S(GTFT,GTFT) = 0, & \quad v_P(GTFT,GTFT) = 0.
\end{align*}

>>>>

\definecolor{darkblue}{rgb}{0,0,.8}
\definecolor{darkgreen}{rgb}{0,0.5,0.1}
\definecolor{lightapricot}{rgb}{0.99, 0.84, 0.69}
\newcommand{\matlabfunc}[1]{\textcolor{darkblue}{#1}}
\newcommand{\matlabcomment}[1]{\textcolor{darkgreen}{#1}}
\newcommand{\christian}[1]{\textcolor{blue}{\textbf{CH}: #1}}
\newcommand{\alex}[1]{\textcolor{red}{\textbf{AL}: #1}}

@@

\title{\bf  \sffamily \LARGE Supplementary Information: Evolution of cooperation among individuals with
limited updating payoff memory\\}
\date{}
\author{Nikoleta E. Glynatsi, Christian Hilbe, Alex McAvoy}

@@

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. The pairwise comparison process consists of three phases;
(1) the mutation phase (2) the game phase and (3) the update phase. In the update
phase an individual adopts the strategy of another individual based on their
``updating payoffs''. In Section~\ref{section:updating_payoffs} we
present/describe the conventional approach for calculating updating payoffs, as
well as, our newly introduced approach.

@@

Pairwise comparison process is a stochastic process for modelling the evolution
of a finite population. The process starts with assigning all individuals of the
population the same strategy. A strategy is a set of rules of how an individual
should behave in an interaction with another individual. Each elementary time
step of the process consists of three phases; (1) the \textbf{mutation phase}
(2) the \textbf{game phase} and (3) the \textbf{update phase}. These are summarised
in Figure~\ref{fig:pairwise_phases}.

In the \textbf{mutation phase} one individual is
chosen to switch to a new mutant strategy with a probability \(\mu\).
In the \textbf{game phase} individuals are randomly matched with other
individuals in the population, and they engage in a repeated game where each
subsequent turn occurs with a fixed probability \(\delta\). At each turn the
individuals decide on an action based on their strategies.
In repeated games there are infinite many strategies, however, it is commonly
assumed that individuals can only choose strategies from a restricted set. One
such set is that of reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently.

In the \textbf{update stage} where two individuals are randomly selected. From
the two individuals, one  serves as the `learner' and the other as the `role
model'. The learner adopts the role model's strategy with a probability \(\rho\)
given by,

\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{RM}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_{RM}\!-\! \pi_{L})}}.
\end{equation}

\(\pi_{L}\) and \(\pi_{RM}\) are the updating payoffs of the learner and the
role model respectively. The updating payoffs are a measure of how successful
individuals are in the current standing of the population. The parameter
\(\beta\) is known as the selection strength, namely, it shows how important the
payoff difference is when the learner is considering adopting the strategy of
the role model.

\begin{figure}[!htbp]
    \centering
    \includestandalone[width=\textwidth]{static/pairwise_comparison_process_diagram}
    \caption{\textbf{Pairwise comparison process phases.} \textbf{A) Initialisation} The process begins
    with a finite population where each member is assigned a given strategy.
    Each color represents a different strategy, and the members are labelled by
    letters. \textbf{B) Mutation phase.} An individual is selected (in the
    example individual A) and with a given probability \(\mu\) that individual
    adopts a new strategy. \textbf{C) Game phase.} Individuals are
    selected to interact in a repeated social dilemma with other individuals. We
    demonstrate the case where individuals B and F have been selected to
    interact. They use the reactive strategies \(s_{B} = (y_B, p_B, q_B)\) and
    \(s_{F} = (y_F, p_F, q_F)\) respectively. The opening moves depend on their
    \(y_i\) probability. In turn 1, individual B cooperated,
    thus, F cooperates with a probability \(p_{F}\) in turn 2. On the opposite,
    individual F defected in turn 1, and so B cooperates in the next turn with
    a probability \(q_{B}\). At each turn there is a probability \(\delta\) that a
    subsequent turn will occur, and with a probability (\(1- \delta\)) the interaction ends.
    \textbf{D) Update phase.} At the updating phase two individuals are chosen;
    one serves as the role of the learner and the other one as the 
    role model. In our example C adopts D's strategy with a probability
    \(\rho(\pi_{C}, \pi_{D})\) where \(\pi_{C}, \pi_{D}\) denote the updating
    payoffs of the individuals.}\label{fig:pairwise_phases}
\end{figure}

This elementary step of the process (mutation, game and update phases) is
repeated for a large number of time steps, and at each time step we record the
state of the population.

@@

\begin{algorithm}[!htbp]
  \SetAlgoLined $N \leftarrow$ population size\; $k \leftarrow 1$\; resident
   $\leftarrow$ starting resident\; \While{ $t <$ maximum number of steps}{mutant
   $\leftarrow$ random: $\{\emptyset \}\ \rightarrow R^n$\; fixation probability
   $\leftarrow \varphi $\; \If{$\varphi >$ random: $i \rightarrow [0,
   1]$}{resident $\leftarrow$ mutant;}} \caption{Evolutionary
   process}\label{algorithm:pairwise_comparison}
\end{algorithm}

Most of the results we present in this work consider the case of low mutation,
however, we have also verified that the main result holds in the case of
high mutation rates. %ToDo Add reference to section with mu != 0.

@@

The updating payoffs depend on the interactions of the individuals at the game
phase. In this work we assume that in the game phase individuals are matched in
pairs and that they participate in a repeated 2 persons donation game. In the
donation game there are two actions: cooperation (\(C\)) and defection (\(D\)).
By cooperating a player provides a benefit \(b\) to the other player at their
cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn are,

Let \(\mathbf{u} = (b-c, -c, b, 0)\) be payoffs in a vector format, and let \(\mathcal{U}
= \{r, s, t, p\}\) denote the set of feasible payoffs, where \(r\) denotes the
payoff of mutual cooperation, \(s\) the sucker's payoff, \(t\) the temptation to
defect payoff, and \(p\) the punishment payoff.

In the following subsections we present several approaches for calculating
the updating payoffs. Initially, we discuss the conventional approach of the
expected payoffs and afterwards we present our newly introduced approach.

@@

The expected payoffs are the conventional payoffs used in the updating stage.
They are defined as the mean payoff of an individual in a well-mixed population
that engages in an infinitely repeated games with all other population members. In
an infinitely repeated game the payoff of a reactive strategy can explicitly be
calculated using a markovian approach~\cite{}. Namely, assume two reactive
strategies $s_1\!=\!(y_1, p_1, q_1$) and $s_2\!=\!(y_2,p_2,q_2)$, their play can
be defined as a Markov process with the transition matrix \(M\),

\begin{equation}\label{eq:transition_matrix}
  \input{static/matrix.tex}.
\end{equation}

Note that \(y_1 \text{ and } y_2\) are omitted since they have no effect in the
long run of the game. The stationary vector \(\mathbf{v}\),
which is the solution to \(\mathbf{v} M = \mathbf{v}\), combined with the payoff vector
\(\mathbf{u}\), yields the game stage outcome for each strategy,

\[\langle\mathbf{v}(s_1,s_2),\mathbf{u}\rangle \quad  \text{ and } \quad  \langle\mathbf{v}(s_2,s_1), \mathbf{u}\rangle.\]

In the case of low mutation there can be only one type of mutant strategy in the
population. So in a population of size \(N\), there will be \(k\) mutants and
\(N - k\) residents, whose strategies we denote respectively as \(s_M =(y_M,
p_M, q_M)\) and \(s_R = (y_R, p_R, q_R)\).
The expected payoffs of a resident (\(\pi_R\)) and for a
mutant (\(\pi_M\)) are give by,

@@

The number of mutants in the population increases if a resident adopts the strategy
of a mutant, and decreases if a mutant adopts the strategy of a resident. The
probabilities that the number of mutants decreases and increases,
\(\lambda^-_k\) and \(\lambda^+_k\), are now explicitly defined as,

\begin{align*}
  \lambda^-_k \!=\!\rho(\pi_M, \pi_R) \quad \text{ and } \quad \lambda^+_k \!=\!\rho(\pi_R, \pi_M).
\end{align*}

\subsubsection*{Simulation Results based on the expected payoffs}

We can simulate the evolutionary process described by
Algorithm~\ref{algorithm:pairwise_comparison} when individuals use reactive
strategies and update their strategies based on their expected payoffs. We
performed two independent runs of the process where we differed the benefit of
cooperation \(b\). The results are shown in
Figure~\ref{fig:expected_payoffs_results}.

It is observed that a higher value of benefit results in a more cooperative
population. For a low benefit the resident population cooperates on average 52\%
of the time. For a a high benefit the average cooperation increases to 98\%. For
both a low and a high benefit the resident population consists either of
defectors or conditional cooperators. A conditional cooperator, otherwise
known as a generous tit for tat (GTFT) strategy, always cooperates if the
co-player cooperated ($p\approx1$) and cooperates with a probability \(q\) if
the co-player defected. The conditional cooperator strategy adopted by the
population differs between the two simulations. A higher value of benefit
results to a higher tolerance to defection.

\begin{figure}[!htbp]
    \centering 
    \includegraphics[width=.70\textwidth]{static/expected_payoffs_donation_game.pdf}
    \caption{\textbf{Evolutionary dynamics under expected payoffs with
    low (left) and high (right) benefit.} We run the simulations for \(T =
    10^7\) time steps. For each time step, we have recorded the current resident
    population \((y, p, q)\). Since in the case of the expected payoffs the
    individuals interact for an infinite number of times \(\delta \rightarrow
    1\), we do not report the players' initial cooperation probability \(y\).
    The graphs show how often the resident population chooses each combination
    \((p, q)\) of conditional cooperation probabilities in the subsequent
    rounds. In both cases players update based on their expected payoffs (A) In
    the case where \(b=3\), the resident population either consists of defectors
    (with \(p \approx q \approx 0\)) or of conditional cooperators for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{3}=0.7\). (B) In the case where
    \(b=10\), the resident population typically applies a strategy for which \(p
    \approx 1\) and \(q \leq 1 - \frac{1}{10}=0.9\). Parameters: \(N =100, c=1,
    \beta=1\).
    }\label{fig:expected_payoffs_results}
\end{figure}

\subsubsection*{Invasion Analysis based on the expected payoffs}

We can explicitly calculate the value of \(q\) which evolves. Namely, we will
calculate how easily a single ALLD mutant can invade into a resident population
with strategy GTFT. In that case, GTFT\(= (1, 1, q)\), ALLD\(= (0, 0, 0)\), and
\(k = 1\).

When two GTFT players interact in an infinitely repeated game, their respective
probabilities for each of the four outcomes in the last round simplify to,

\begin{align*}
    \mathbf{v}(\text{GTFT}, \text{ALLD}) = (1, 0, 0, 0).
\end{align*}

@@

\begin{align*}
  \mathbf{v}(\text{ALLD}, \text{GTFT}) = (0, q, 0, (1 - q)).
\end{align*}

@@

\begin{align*}
  \displaystyle \pi_{\text{GTFT}} & = \displaystyle \frac{N\!-\!2}{N-1} (b - c)  -	\displaystyle\frac{q c}{N-1} \quad \text{ and } \quad \displaystyle \pi_{\text{ALLD}}  = \displaystyle b q.
\end{align*}

@@

\begin{align*}
    \frac{\lambda^{+}}{\lambda^{-}} & = \frac{\rho(\pi_{\text{GTFT}}, \pi_{\text{ALLD}})}{\rho(\pi_{\text{ALLD}}, \pi_{\text{GTFT}})}  = \frac{e^{-\beta \left(\frac{(N-2) (b-c)}{N-1} - q (b + \frac{c}{N-1})\right)}+1}
    {e^{-\beta\left(b q - \frac{b (N - 2)}{N-1} - \frac{c (N - 2 - q)}{N-1}\right)}+1}
\end{align*}

@@

\subsection{Updating Payoffs based on the last last \(m\) round(s) payoffs of \(n\) interaction(s)}

The aim of this work is to explore the effect that the updating payoffs have on
the cooperation rate of the evolved population. To this end we introduce
a new method calculating the updating payoffs. Compared to the literature
where a infinite number of rounds and a well mixed population are assumed,
we constrain the number of turns and number of turns.

Namely, in the expected payoffs the payoff of a pair depends on the average
payoff they received over an infinite number of turns. In our approach, the
payoff of a pair depends on the average payoffs they received in the last \(m\)
turns. Furthermore, in expected payoffs it is assumed that a player interacts
with every member of the population, whereas in our approach with consider one
\(n\) interactions of the player with other members. Here we present results for
four cases; (\(m=1\) and \(n=1\)), (\(m=2\) and \(n=1\)), (\(m=1\) and \(n=2\)),
and (\(m=2\) and \(n=2\)).

\subsubsection{Updating Payoffs based on the last round payoff of one interaction (\(m=1\) and \(n=1\))}\label{section:m_one_n_one}

In this case an individual updates her/his strategy based on the last payoff
they received against one other member of the population. Initially, we define
the probability that a reactive strategy receives the payoff $u\!\in\!
\mathcal{U}$ in the very last round of the game against another reactive
strategy (Proposition~\ref{proposition:last_round}).

@@

      v_{r}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1y_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(q_1+l_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(q_2+l_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
      {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times r,\\[1cm]

      v_{s}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1\bar{y}_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(q_1+l_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
      {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times s,\\[1cm]

      v_{t}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1y_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(q_2+l_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
      {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times t,\\[1cm]

      v_{p}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1\bar{y}_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
      {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times p.
      \end{array}
    \end{equation}

In these expressions, we have used the notation $l_i:=p_i\!-\!q_i$,
$\bar{y}_i\!=\!1\!-\!y_i$, $\bar{q}_i:=1\!-\!q_i$, and
$\bar{l}_i:=\bar{p}_i\!-\!\bar{q}_i=-l_i$ for $i\!\in\!\{1,2\}$.
\end{Prop}

Note that in the proposition we here we focus on the case of the donation game/prisoner's
dilemma but the result applies to any \(2 \times 2\) symmetric game.

@@

At each step of the evolutionary process we choose a role model and a learner to
update the population. We consider the case where both the role model and the
learner estimate their fitness after interacting with a single member of the
population. There are five possible pairings at each step. They interact with
other with a probability \(\frac{1}{N - 1}\), and thus they do not interact with
other with a probability \(1 - \frac{1}{N - 1}\). In the latter case, each of
them can interact with either a mutant or a resident. Both of them interact with
a mutant with a probability $\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact
with a resident with a probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last
two possible pairings are that either of them interacts with a resident whilst
the other interacts with a mutant, and this happens with a probability
$\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$.

Given the possible pairings and Proposition~\ref{proposition:last_round}, we
define the probability that the respective last round payoffs of two players
\(s_1, s_2\) are given by $u_1$ and $u_2$ as,

@@

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the game stage, which happens with
probability $\frac{1}{(N\!-\!1)}$. In that case, we note that only those payoff
pairs can occur that are feasible in a direct interaction,
$\splitatcommas{(u_1,u_2)\in \mathcal{U}^2_F:=\big\{ (r, r), (s, t), (t, s), (p, p)
\big\}}$, as represented by the respective indicator function. Otherwise, if the
learner and the role model did not interact directly, we need to distinguish
four different cases, depending on whether the learner was matched with a
resident or a mutant, and depending on whether the role model was matched with a
resident or a mutant.

The probability that the number of mutants increases, and decreases respectively,
by one is now given by,

@@

\subsubsection*{Simulation Results based on the last round payoff of one interaction}

We simulate the evolutionary process given that individuals now use the the last
round payoff of one interaction to update their strategies. The results are
shown in Figure~\ref{fig:one_interaction_last_round_payoffs_results}. Similarly
to the results of the expected payoffs, a higher benefit results in a more
cooperative population. However we note that the cooperation rate of the evolved
populations are lower compared to the case of expected payoffs, and that the
cooperation rate increases less between the two simulations. The evolving both
consist of defectors and conditional cooperators. The evolved \(q\)s are smaller
which also results to a less cooperative population. In between the two
simulations \(q \approx \frac{1}{2}\). In the invasion analysis we will
demonstrate that in the case of the the last round payoff of one interaction, a
conditional player needs to have \(q \leq \frac{1}{2}\) to avoid being invaded
by a defector. In both simulations \(q \leq \frac{1}{2}\), the higher benefit
pushes the evolved population closer to the bounty.
